---
title: 'Wrangling data with the tidyverse'
subtitle: '2031 - Introduction to Statistical Computing'
author: 'Sam Clifford'
date: '2020-11-19'
output:
  beamer_presentation:
    keep_tex: true
    theme: 'Hannover'
    colortheme: 'dove'
    latex_engine: xelatex
    citation_package: natbib
    includes: 
      in_header: matrix.tex
header-includes:
 - \usepackage{longtable}
bibliography: extras.bib
csl: chicago-author-date.csl
---

```{r global_options, include=FALSE, cache=FALSE}
library(tidyverse)
library(ggthemes)
library(knitr)

opts_knit$get('out.format')


#output <- opts_knit$get('rmarkdown.pandoc.to')
knitr::opts_chunk$set(echo = FALSE, collapse=T, warning=FALSE, message=FALSE,
                      fig.align='center', out.width = '100%',
                      fig.width=4,  fig.height=2)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

old_patterns = knit_patterns$get() # old pattern list (to restore later)
all_patterns = all_patterns # a list of all built-in patterns
knit_patterns$set(all_patterns[["rnw"]]) # set pattern list from 

```


# Introduction

### Tidyverse

* The *tidyverse* suite of R packages is designed to make working with data as easy as possible [@tidyverse]
* Focus is on operating on objects that behave like data frames
* The relevant packages from tidyverse for us are
    - ggplot2: for plotting data (next week)
    - dplyr: for manipulating data frames
    - tidyr: for making data tidy
    - forcats: for manipulating factor variables
    - *magrittr: for easy chaining of commands*
    
[@r4ds; @welcometidyverse]

\small

``` {r eval=FALSE, echo=TRUE}
library(tidyverse)
```

# Reshaping data frames

### Long and wide tidy data

* Every data set has its own quirks 

<!-- * Every modelling package has its own way of returning outputs -->
<!--     - data frame -->
<!--     - matrix -->
<!--     - list of the above -->

* Tidy **data frames** consist of a number of observations (rows) of variables (columns), they can be either **wide** or **long** 

* Data usually stored as
    * 1 row per observations
    * 1 column per variable
    * $X_{ij}$ is $i$th observation of $j$th variable
    * Tidy because all observations have an entry for each variable (even if it's NA)

* Data needs to be the right shape for the functions being used

[@JSSv059i10]

<!-- * ggplot2 (next week) usually requires long data -->


### Long and wide tidy data

* An example of a wide data frame which we might encounter is the output of an SIR model

\small

``` {r echo=FALSE}
library(tidyverse)
SIR <- read_csv('beta_1.4247_gamma_0.14286.csv')
```

:::::::::::::: {.columns}

::: {.column}

**Wide data **

``` {r}
SIR %>% 
  filter(time %% 0.5 == 0, time <= 1) %>%
  knitr::kable(., 
               digits=6,
               format = 'latex',
               booktabs = TRUE)
```


* *key*: this `state` at this `time`
* *value*: `proportion`


:::

::: {.column}


**Long data **

``` {r}
SIR_long <- 
  pivot_longer(data            = SIR, 
               cols            = c(S, I, R), 
               names_to        = 'state', 
               values_to       = 'proportion')

SIR_long %>% 
  filter(time %% 0.5 == 0, time <= 1) %>%
  knitr::kable(., 
               digits=6,
               format = 'latex',
               booktabs = TRUE, linesep='')
```

:::
::::::::::::::


<!-- \normalsize -->

<!-- * Here we have values of $S(t), I(t), R(t)$ at given values of $t$ -->


### Long and wide tidy data

* Our numerical solution to the SIR model is a wide data frame, values of $S(t), I(t), R(t)$ at given values of $t$ 

* We *pivot* the columns in `SIR` so that the data frame is *longer*
* This pivoting to a longer data frame helps us put the data in *key*-*value* pairs
* The key is the unique identifier
    * state - $S$, $I$, or $R$, and
    * time
* The value is the proportion of the population in this state at this time

### Long and wide tidy data

* To make this pivot, we specify in tidyr's `pivot_longer()` function
    * which `cols` are to be converted from being $k$ columns of length $n$ to one column of length $n \times k$
    * the *names* column, `state`, contains the names of the columns being pivoted
    * the name of the column containing the *value* (proportion) of each `state` at given `time`


\footnotesize

``` {r gather_sir, echo=TRUE}
SIR_long <- pivot_longer(
  data            = SIR, 
  cols            = c(S, I, R), 
  names_to        = 'state', 
  values_to       = 'proportion')
```

* NB: we need to use `'` quotes for `names_to` and `values_to` arguments because they are strings defining new columns

### Long and wide tidy data


:::::::::::::: {.columns}
::: {.column}

**Wide data **

``` {r}
SIR %>% 
  filter(time %% 0.5 == 0, time <= 1) %>%
  knitr::kable(., 
               digits=6,
               format = 'latex',
               booktabs = TRUE)
```



* *key*: this `state` at this `time`
* *value*: `proportion`

:::
::: {.column}


**Long data **

``` {r}
SIR_long <- pivot_longer(
  data            = SIR, 
  cols            = c(S, I, R), 
  names_to        = 'state', 
  values_to       = 'proportion')

SIR_long %>% 
  filter(time %% 0.5 == 0, time <= 1) %>%
  knitr::kable(., 
               digits=6,
               format = 'latex',
               booktabs = TRUE, linesep='')
```

:::
::::::::::::::

### Long and wide tidy data

* To convert to a wider format, we use `pivot_wider`
* For example, we specify:
    - the *data* source
    - where we get the new column *names from* 
    - where we get the new column *values from*

``` {r, echo = TRUE, size = "scriptsize"}

SIR_wide <- pivot_wider(
  data        = SIR_long, 
  names_from  = state,
  #NB: these are variable names, so doesn't require ' quotes
  values_from = proportion)

head(SIR_wide, 3)

```

* We have recovered our original simulation data frame

# Summarising data

### Summarising data

\small

* One of the most common data tasks is calculating summary statistics
* Consider the `Gestation` data from `mosaicData`

``` {r}
library(mosaicData)
data(Gestation)
```

> Birth weight, date, and gestational period collected as part of the Child Health and Development Studies in 1961 and 1962. Information about the baby's parents — age, education, height, weight, and whether the mother smoked is also recorded.

* We will use some functions from dplyr to choose, group and summarise data

### Summarising data

* `count` how many mothers in each level of the `race` variable

``` {r, echo = TRUE, size = "small"}
count(Gestation, race)
```

* Calculate the mean birth weight in the data set

``` {r, echo = TRUE, size = "small"}
summarise(Gestation, Wt = mean(wt))
```

### Summarising data

* We can calculate multiple summaries at once

``` {r, echo = TRUE, size = "scriptsize"}
summarise(Gestation,
          Mean = mean(wt),
          SD   = sd(wt),
          Low  = quantile(wt, 0.025),
          High = quantile(wt, 0.975))
```

or summarise multiple variables with the same function

``` {r, echo = TRUE, size = "scriptsize"}
summarise_at(Gestation, 
             .vars = vars(gestation, wt, age),
             .funs = list(mean), na.rm = T)
```

### Summarising data

* We can even do multiple summaries across multiple variables by specifying which `.vars` we want to summarise and with which `.funs` we wish to summarise with

``` {r, echo = TRUE, size = "scriptsize"}
summarise_at(Gestation, 
             .vars = vars(gestation, wt, age),
             .funs = list(mean = mean,
                          sd   = sd), na.rm = T)
```

### Summarising data

* Where this is most powerful is when we specify a grouping structure in our data
* `group_by` allows us to pass variable names to set the structure without changing row order
* Result is a tibble with specified grouping structure

``` {r, echo=TRUE, size = "tiny"}
Gestation_grouped_by_race <-
  group_by(Gestation, race)

Gestation_grouped_by_race
```

### Summarising data

* `summarise` respects the grouping structure

``` {r, echo=TRUE, size = "small"}
summarise_at(Gestation_grouped_by_race, 
             .vars = vars(gestation, wt, age),
             .funs = list(mean), na.rm = T)

```

<!-- * N.B. the `NA` race rows have been dropped because they don't have a grouping level -->

# Relevelling factors

### Relevelling factors

* Default behaviours are:
    - `pivot_longer()` respects column order when reshaping
    - `key` column is character variable
    - character variables coerced to alphabetic factors
* We can set order of a factor variable by specifying levels
* forcats package contains many `fct_*` functions to help with this
* dplyr's `mutate()` to create, modify or delete a column in our data frame


### Relevelling factors

\small

* Consider the `Gestation` data from `mosaicData`
* May not make sense to sort race alphabetically
* `NA` is not descriptive
* Let's replace `NA` with an explicit factor level and sort differently

``` {r echo=TRUE, size = "scriptsize"}
Gestation <- mutate(
  Gestation,
  race2 = factor(race), 
  race2 = fct_explicit_na(f        = race2, 
                          na_level = "unknown"))

count(Gestation, race2)
```

### Relevelling factors

\small

* This is now slightly more human friendly
* We may wish to set the levels in order of how frequent they are

``` {r echo=TRUE}
Gestation <- mutate(
  Gestation,
  race2 = fct_infreq(race2))

count(Gestation, race2)
```

### Relevelling factors

* Many other functions in forcats
    * `fct_inorder()` - for creating a factor variable with levels in the order from top to bottom
    * `fct_relevel()` - change the order of levels
    * `fct_recode()` - change the value of factor levels
    * `fct_rev()` - reverse the existing levels of a factor
    * `fct_relabel()` - relabel levels according to some function

# Choosing rows and columns

### Choosing columns

* For one reason or another we may want to `select` only certain columns of our data frame

``` {r, echo=TRUE, size = "tiny"}
head(Gestation)

head(select(Gestation, race2, wt, number))
```

### Choosing and renaming columns

\small

* We can also rename columns on the fly as we select them

``` {r, size = "tiny", echo = TRUE}
select(Gestation, Race     = race2, 
       `Birthweight (lbs)` = wt,
       `Cigs. smoked`      = number)
```

* Alternatively we can just `rename` columns without worrying about failing to select what we want to keep

``` {r, size = "scriptsize", echo=TRUE}
Gestation <- rename(Gestation, `Cigs. smoked` = number)
```

### Choosing rows

* The dplyr equivalent of `subset` is `filter`
* Takes a logical statement and does non-standard evaluation of variable names
    - can use `filter(data, A & B)` or `filter(data, A, B)` equivalently

``` {r echo=TRUE, size = "scriptsize"}
Gestation2 <- select(Gestation,
                     Race                = race2, 
                     `Birthweight (lbs)` = wt,
                     `Cigs. smoked`)

filter(Gestation2, 
       Race == "white", `Cigs. smoked` == "never")
```

### Arranging rows

* The `arrange` function rearranges rows in ascending order of specified variables
* Can be useful for exploratory analysis or if you need to create a column based on ranking of data


``` {r, echo=TRUE, size = "tiny"}
arrange(Gestation, date, wt)

```

# Pipe

### Pipe

\small

* dplyr imports the `%>%` pipe from magrittr
* `f(g(x))` is equivalent to `x %>% g %>% f`
* Makes it easier to chain operations together without storing temporary objects
* Output on left of `%>%` becomes first argument of function on right
    - by convention, all tidyverse functions take a data frame as their first argument

``` {r, eval = FALSE, size = "scriptsize", echo=TRUE}

x %>% f_1 %>% f_2 %>% f_3

# rather than

f_3(f_2(f_1(x)))

# or

x_1 <- f_1(x)
x_2 <- f_2(x_1)
x_3 <- f_3(x_2)
```

### Pipe

* An example

``` {r echo, size = 'tiny', echo=TRUE}
Gestation %>%
  group_by(race2, `Cigs. smoked`) %>%
  summarise(wt = mean(wt)) %>%
  pivot_wider(names_from  = race2,
              values_from = wt)

```


* Some find this a little confusing, so we won't dwell on it
* NB the `Cigs. smoked` variable is character and sorted alphabetically.

# Summary

### Summary

* Reshaping data frames
    - `pivot_longer` from variables side by side to key-value
    - `pivot_wider` from key-value to named column variables
* Summarising data
    - `group_by` to set a group structure
    - `summarise` to calculate summary stats across group structure
    - `count` to see how many rows in each group
* Relevelling factors
    - `factor` to set a factor variable
    - `fct_*` for modifying factors
* Choosing rows and columns
    - `select` to choose columns
    - `filter` to choose rows
    - `arrange` to sort rows
    
### Additional Resources

* More help on the [tidyverse](https://www.rstudio.com/resources/cheatsheets/) is available
* The `#r4ds` community have [TidyTuesday](https://github.com/rfordatascience/tidytuesday) which makes use of the ideas in the R for Data Science book [@r4ds]
* @JSSv059i10 on what tidy data is
* @welcometidyverse for an explanation as to what the tidyverse is


