---
title: 'Publication ready graphics'
subtitle: '2941 - Data Challenge'
author: 'Sam Clifford'
date: '2021-02-05'
output: 
  pdf_document:
    includes:
      in_header: matrix.tex
citation_package: natbib
csl: chicago-author-date.csl
bibliography: pubready.bib
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, eval=FALSE,
                      fig.align='center', dpi = 300)
library(tidyverse)

library(extrafont)
loadfonts()

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Introduction

## About this practical session

In the week 1 lecture session on exploratory data analysis we introduced visualisation with the histogram, $x$-$y$ plots and other scatter plot techniques, and touched on Tufte's principles of graphical excellence.

This practical will investigate what makes a good and a bad graph, in order to help you generate publication quality graphics. You will first critique a plot produced through the provided code. Then, you may choose which order to do activities 2a and 2b, which are to produce a better and a worse version of the given plot. You should aim to finish both of these during the time given.

* Assumed skills
    - Writing R code into a script file
    - Reading and writing ggplot2 code
    - Identifying things that are visually pleasing
* Learning objectives
    - Identifying the link between code and the graph it produces
    - Being able to critique a graph
    - Understanding why and how data is encoded and decoded visually
    - Understanding the subjectivity of what is aesthetically pleasing
* Professional skills
    - Creating high quality graphics

## Group formation

You will be allocated to groups of 3-4 in Zoom breakout rooms.

<!-- Assign the following roles. Someone may need to take on multiple roles or you may need to split a role depending on the size of your group. -->

``` {r, echo=FALSE, results='asis', fig.align='center', eval=FALSE, fig.width=7, fig.height=5, fig.align = 'center', out.width = '100%'}
library(pander)
panderOptions('table.alignment.default', 'center')
data.frame(Role = c('Chair', 
                    'Scribe', 
                    #'Research', 
                    #'Programmer',
                    #'Critical Friend',
                    'Spokesperson'),
           Duties = c('Ensures the group is getting the work done in a timely manner and everyone gets a chance to contribute',
                      'Responsible for recording answers in this worksheet',
                      #'Has the slides ready and able to fetch other relevant information online',
                      #'Drives R with the support of the group',
                      #'Constructive criticism, is the group answering the question?',
                      'Presentation of discussion points to class')) %>%
  mutate(Name = rep('Student Name', nrow(.)),
         Email = rep('Student Email', nrow(.))) %>%
  pandoc.table(justify=rep('left', ncol(.)),
               graph.boxes=TRUE, table.split.table= Inf,
               style='multiline', keep.line.breaks = TRUE,
               split.cells = c('20%', '30%', '20%', '30%'))

```

A reminder of expectations in the prac:

* Keep a record of the work being completed with a well-commented R script
* Allow everyone a chance to participate in the learning activities, keeping disruption of other students to a minimum while still allowing for fruitful discussion
* All opinions are valued provided they do not harm others
* Everyone is expected to do the work, learning seldom occurs solely by watching someone else do work

# Activity 1 - Building an attempt at a plot

We will be looking at the gapminder data set as found in the gapminder package [@gapminder]. This data has been collected from countries around the world and contains data on life expectancy, population and GDP per capita for 142 countries from 1952 to 2007.

**Exercise:** Copy and paste the code below to produce a plot showing how the relationship between GDP, life expectancy and population vary over time and continent. If you can't install the gapminder package, you can download the data from Moodle and load it with `read_csv()` from the readr package (loaded when tidyverse is loaded). Save the plot as both a PDF and PNG and zoom in to see the pixellation in the PNG image.

``` {r, eval=TRUE, fig.width = 8, fig.height = 4, size = "small"}
library(gapminder)
library(tidyverse)
data(gapminder)

library(extrafont)
loadfonts()

gapminder_example <- 
  ggplot(data = gapminder,
         aes(x = gdpPercap, y = lifeExp)) +
  geom_path(aes(group = country, color = continent)) +
  geom_point(aes(color = continent, size = pop)) +
  scale_color_brewer(palette = 'Set2') + 
  scale_x_log10() +
  annotation_logticks(sides = 'b') +
  facet_wrap( ~ continent, scales = 'free', nrow = 1) +
  scale_size_area() +
  theme_dark() +
  labs(x = "GDP", y = "LE",
       title = "GDP and Life Expectancy 1952-2007") +
  theme(legend.position = 'bottom',
        legend.box      = 'vertical',
        axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(family = "Times New Roman"))

gapminder_example

ggsave(filename = "gapminder_example.pdf",
       plot = gapminder_example, width = 8, height = 4, 
       units = "in")


ggsave(filename = "gapminder_example.png",
       plot = gapminder_example, width = 8, height = 4, 
       units = "in", dpi = 300)

```

\normalsize

**Exercise:** Discuss, within your group, what you think is good and bad about this plot. Does it conform to Tufte's principles of graphical excellence? Is it easy to interpret? Does it show the relationship we are interested in? List *three* important improvements that are needed for this graph to be useful. This should take no longer than ten minutes.

<!-- **Answer:** The plot shows each country's data connected by a line so it's clear that individual country's GDP and life expectancy changes over time in a sequence, although we don't know which direction along the line we are travelling forward in time. By separating each continent out, we can see the trends geographically, such as Europe being clustered quite closely and Africa being very scattered (lots of variation across space and time, rather than a nice orderly procession from bottom left to top right). Logarithmic axis helps put the lines at approximately 45 degree angles rather than as difficult to read logarithmic curves. The large points obscure the data. Continent is mapped to both colour and facet. It's hard to make comparisons across continents because the $x$ and $y$ scales are different, so we have to think very hard about whether Asia is richer or poorer than Europe, on average, and we may miss that there is very little overlap in the Oceania and Africa values. -->

<!-- Three things which might be worth changing: -->

<!-- 1. Same axis scaling -->
<!-- 2. Show less data (either fewer variables or fewer continents/years) -->
<!-- 3. Put time on the $x$ axis -->

**Exercise:** As a group, discuss what you think each line of code in the above block does. You may wish to answer as comments in your code (everything after a `#` is a comment) or in a separate document.

<!-- **Answer:** -->

<!-- ``` {r, eval=FALSE, fig.width = 8, fig.height = 3} -->
<!-- library(gapminder) # load package with data -->
<!-- library(tidyverse) # load package to manipulate and visualise data -->
<!-- data(gapminder)    # load the gapminder data -->

<!-- # make a plot with the gapminder data -->
<!-- ggplot(data = gapminder, -->
<!--        # put GDP as X variable and LE as Y -->
<!--        aes(x = gdpPercap, y = lifeExp)) + -->
<!--   # draw the X-Y pairs as a line, ordered by appearance in data frame -->
<!--   # group the lines by country and colour them by the continent -->
<!--   geom_path(aes(group = country, color = continent)) + -->
<!--   # draw points at each X-Y pair, coloured by continent and  -->
<!--   # with their size based on the POP variable -->
<!--   geom_point(aes(color = continent, size = pop)) + -->
<!--   # change the colour scheme from the default for any colour aesthetics -->
<!--   # make a logarithmic scale on the X axis -->
<!--   scale_color_brewer(palette = 'Dark2') + scale_x_log10() + -->
<!--   # put logarithmic ticks on the bottom X axis showing that we don't have -->
<!--   # a uniform scaling -->
<!--   annotation_logticks(sides = 'b') +  -->
<!--   # put each continent on its own set of axes, with all plots in one row -->
<!--   facet_wrap( ~ continent, scales = 'free', nrow = 1) + -->
<!--   # the size of the points, based on the POP variable, should have an area -->
<!--   # proportional to value of POP, rather than their radius -->
<!--   scale_size_area() + -->
<!--   # put the plot legend at the bottom and stack the variable keys vertically -->
<!--   theme(legend.position = 'bottom', -->
<!--         legend.box = 'vertical') -->

<!-- ``` -->

# Activity 2a -- Making a better graph

Based on the ideas discussed, build a graph which your group believes better shows the relationship between life expectancy and GDP. Think first about what story you want your plot to tell; are you interested in trends over space and/or time? Are you interested in a particular continent or even just one country? Do you want to show a snapshot in a particular year, or compare across two or more years?

You may choose to either modify the code given above or create your own graph from scratch. Make sure your code is written in your script file with appropriate comments.

Some things you may wish to consider: 

* meaningul annotations
* overall theme
* what is the most appropriate geometry to show your relationship of interest?
* what aesthetics [Figure 5, @Kunz2011] can help us highlight key context?

You may wish to sketch the graph by hand before attempting to write the R code to generate it. This may help you and your group come to an agreement about the plot you want to make and will help the tutors understand what you're aiming for when you ask them for help.

If you get stuck, look at the following resources for help

* [ggplot2 documentation](https://ggplot2.tidyverse.org/reference/)
* [R Graphics Cookbook](https://r-graphics.org/) [@changgraphics]
* [Chapter 3 and 4 of R for Data Science](https://r4ds.had.co.nz/) [@r4ds]
* [RStudio cheatsheets](https://www.rstudio.com/resources/cheatsheets/) [@cheatsheets]

**Exercise:** Make a plot, save it to your computer and write comments in your code or standalone document that outline what the changes you made were and why.

<!-- **Answer:** There are many ways to do this, but a scatter plot which is static in time and shows the variability across space can help us avoid showing too much. Essentially we run out of things to reasonably change to show all dimensions of the data. -->

<!-- 1. Reduce number of variables shown -->
<!-- 2. Remove faceting, relying on colour to show difference -->
<!-- 3. Clarity around units of variables -->
<!-- 4. Human friendly axis labels -->

``` {r, fig.height = 5, fig.width = 7, eval=FALSE, echo=FALSE}
p <- ggplot(data = filter(gapminder, year == 2007),
       aes(x = gdpPercap/1000, y = lifeExp)) +
  geom_point(aes(group  = country,#
                 size   = pop*1e-6,
                 color  = continent),
             alpha = 0.75) +
  scale_color_brewer(palette = "Dark2", name = "Continent") +
  scale_x_log10() +
  scale_size_area(name = "Population (millions)") +
  theme_bw() + theme(legend.position = "bottom", legend.box = "vertical") +
  xlab("GDP per capita, adjusted for inflation (1000 USD)") +
  ylab("Life expectancy at birth (years)") +
    theme(panel.grid.minor.x = element_blank()) +
    annotation_logticks(sides = "bt") +
  ggtitle("Global Health and Wealth, 2007")

p 

```

<!-- Another solution is -->

``` {r, fig.height = 5, fig.width = 7, eval=FALSE, echo=FALSE}
library(countrycode)
library(ggthemes)
p_big <- filter(gapminder, year %in% c(1997, 2007)) %>%
  mutate(iso3n = countrycode(sourcevar = country,
                             origin = "country.name",
                             destination = "iso3n", origin_regex = T)) %>%
  mutate(
    continent = countrycode(iso3n, "iso3n", destination = "un.region.name"),
    region    = countrycode(iso3n, "iso3n", destination = "un.regionsub.name"),
    region    = ifelse(iso3n == 158, "Eastern Asia", region),
    continent = ifelse(iso3n == 158, "Asia", continent),
    region    = fct_collapse(region,
                             `CANZUS` = c("Northern America",
                                          "Australia and New Zealand"))) %>%
  arrange(continent, region) %>%
  mutate(region = fct_inorder(region)) %>%
  ggplot(data = .,
         aes(x = gdpPercap/1000, y = lifeExp)) +
  geom_line(aes(group = country,
                color = continent),
            arrow = arrow(type = 'closed',
                          length = unit(0.1, "cm"))) +
  facet_wrap(~region, ncol = 4) +
  scale_x_log10(labels = function(x){sprintf("%g",x)}) +
  theme_bw() +
  theme(legend.position  = "bottom",
        legend.box       = "vertical",
        panel.grid.minor = element_blank(),
        text = element_text(family = "Roboto Condensed"),
        axis.title.x = element_text(hjust = 1)) +
  xlab("GDP per capita, adjusted for inflation (1000 USD)") +
  ylab("Life expectancy at birth (years)") +
  annotation_logticks(sides = "bt") +
    scale_colour_brewer(palette = "Dark2", name = "Continent") +
    ggtitle(label    = "National GDP and life expectancy at birth",
            subtitle = "Change from 1997-2007")


```

# Activity 2b -- Making a worse graph

Make a new graph as in the previous activity but make it as bad as possible while still attempting to honestly show the information (i.e. don't add things to the plot which can't be derived from the variables in the plot).  Your plot should be an honest attempt to show the data poorly, rather than a deliberately unreadable mess. Think of something you'd expect to see in a newspaper staffed with well-intentioned but unskilled staff.

Consider the principles of graphical excellence and how can we go against them to make a terrible plot. Think about what was bad about the plot provided earlier. Consider abusing the ability to map graphical options (e.g. color, fill, line type, point size) to our variables of interest.

**Exercise:** Make a plot, save it to your computer and write comments in your code or standalone document that outline what the changes you made were and why.

<!-- **Answer:** While the graph is already quite bad, we can make it worse by removing the focus on the relationship between GDP and LE, shifting one of them out of the $x$ and $y$ variables. -->

``` {r, fig.height = 5, fig.width = 7, eval=FALSE, echo = FALSE}

fix_cut_labels <- function(x){
  levs <- levels(x)
  
  new_levs <- 
    strsplit(levs, split = ",") %>%
    map(parse_number) %>%
    map_chr(~paste0("(",paste(.x, collapse = ", "), "]"))
  
  fct_recode(x, !!!set_names(levs, new_levs))
  
}


gap <- gapminder %>%
  group_by(continent, year) %>%
  summarise(GDP     = weighted.mean(gdpPercap,pop),
            lifeExp = weighted.mean(lifeExp, pop)) %>%
  ungroup %>%
  mutate(GDP_ = cut(GDP,
                          breaks = 10^seq(2, 5, by = 1))) %>%
  mutate(GDP_ = cut(GDP, breaks = 8)) %>%
  mutate(GDP_ = fix_cut_labels(GDP_)) 

p_bad <- gap %>%
  ggplot(data = ,
         aes(x = factor(year), y = lifeExp)) +
  geom_col(aes(fill = GDP_),
           color = "black") +
  #facet_wrap(~country, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90),
        title = element_text(family = "Copperplate"), 
        plot.subtitle = element_text(color = "grey70"),
        strip.text = element_text(family = "Copperplate", hjust = 1, color = "grey70"),
        panel.grid = element_blank()) +
  scale_fill_brewer(palette = "Greens", name = "Mean GDP ($)") +
  ylab("Mean life expectancy (Y)") +
  xlab("Time (Y)") +
  facet_wrap(~continent, ncol = 1) +
  theme(panel.grid.major.y = element_line(linetype  = 2, color = "grey70")) +
  ggtitle("Life expectancy and GDP in each continent 1952-2007",
          subtitle = "Population-weighted means calculated for each year and continent")

p_bad
```

Here we've made the following changes:

1. Rotated the $x$ axis tick labels too far
2. Coloured bars by GDP, requiring us to mentally map the relationship between bar colour and bar height
3. Let `cut()` choose cut points for GDP that are meaningless (evenly spaced across range)
4. Unnecessary level of precision in GDP
5. Choice of Copperplate font doesn't aid readability
    - moreover, the font families don't match well across the font
6. Put life expectancy's axis label as "Y", same as for time
    - year isn't actually measured in years the same way life expectancy is
7. Many countries use dollars, which dollars do we mean?
8. No indication of variability within continent and year for GDP or life expectancy, both of which have been averaged (weighted by population)
9. Sequential colour scheme is good, but has too many divisions to be useful.
10. Annotations are too pale (subtitle, horizontal grid lines, and facet headings for each continent)
11. Bar plots *must* go to zero (and they do here) but because the values of life expectancy go from 40-80 we find it difficult to tease out trends in life expectancy

# Activity 3 -- Group discussion

Have participants present their best and/or worst graph from the last activities. What did they identify as good and bad and how has each group attempted to present the relationship?

# Tidy up

Make sure you save your R script, and anything else you have produced and ensure everyone in your group has a copy. Email your best and worst graphs to [Dr Sam Clifford](mailto:sam.clifford@lshtm.ac.uk).

# Further reading

Key ideas introduced by @tufte are summarised by @pantoliano. Some of the history of data visualisation is summarised well by @Friendly:05:gfkl and @Friendly:06:hbook. [Tufte's website](https://www.edwardtufte.com/tufte/) is well worth exploring, particularly the discussion on how the visual presentation of information could have helped avert the *Challenger* disaster [@visstatthink1997].

# References

\footnotesize