---
title: 'Exploratory data analysis with ggplot2'
subtitle: '2491 - Data challenge'
author: 'Sam Clifford'
date: '2021-01-14'
output:
  beamer_presentation:
    keep_tex: true
    theme: 'Hannover'
    colortheme: 'dove'
    latex_engine: xelatex
    citation_package: natbib
    includes: 
      in_header: matrix.tex
bibliography: extras.bib
#csl: chicago-author-date.csl
---
  
```{r global_options, include=FALSE}
library(knitr)
opts_knit$get('out.format')
knit_patterns$set(all_patterns[['rnw']])

library(tidyverse)
library(ggthemes)

output <- opts_knit$get('rmarkdown.pandoc.to')
knitr::opts_chunk$set(echo = FALSE, collapse=T, warning=FALSE, message=FALSE,
                      fig.align='center', out.width = '100%',
                      fig.width=4,  fig.height=2)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

```

# What is exploratory data analysis?

### What is exploratory data analysis?

> Since the aim of exploratory data analysis is to learn what seems to be, it should be no surprise that pictures play a vital role in doing it well. There is nothing better than a picture for making you think of questions you had forgotten to ask (even mentally).

@tukeytukey85

### What is exploratory data analysis?

* Combination of graphical and numerical summaries of data [@tukey1972some; @tukey1977exploratory]
* Want to explore relationships in data with view to
    - checking our assumptions about the data 
    - model formulation
    - designing further experiments 
    
### Why do we visualise?

* We don't just make graphs because it's fun (but it totally is!)
* We do it to communicate information
* Describing with summary statistics may not tell the whole story
* Reader should be able to understand what the graph means and not be
    - misled into thinking something that is untrue
    - distracted from the main point


### Tufte's principles

@tufte and @pantoliano

* Show the data
    - substance of results over method details or maker's skills
    - avoid distorting data's meaning
    - serve clear purpose:  description, exploration, tabulation, or decoration

* Provide clarity
    - many numbers, easily digestible
    - coherent representation of large data set
    - closely integrated with statistical and written descriptions of the data set
  
### Tufte's principles

* Allow comparison where appropriate
    - use aesthetics to draw attention to important details
    - make clear that data has multiple levels of structure

* Produce graphs with high data density
    * make every drop of ink count
    * careful use of whitespace
* Avoid excessive and unnecessary use of graphical effects (chartjunk)
* Don't use the same variable for multiple graphical elements (e.g. color and $y$ value)

    
# Building plots

### Building plots

* R package `ggplot2` uses a grammar of graphics [@layered-grammar; @cheatsheets]
    - map variables in data frame to aesthetic options in the plot
    - choose a geometry for how to display these variables
    - adjustments to axis scales
    - adjustments to colors, themes, etc.
    - adding extra commands in a 'do this, then do this' manner
* python users have plotnine [@plotnine] which is based on the same ideas
  
### Building plots

* How do we structure a call to `ggplot` to make a plot?
    - load ggplot2 package
    - Specify we want a `ggplot` object and which data frame we're going to use, 
    - set **aesthetic options** to tell R which variables to map to the $x$ and $y$ axes of the plot
    - state geometry we're using to show variables
    
``` {r, eval=F, echo=T}
library(ggplot2)
ggplot(data = my.data.frame, 
       aes(x = my.x.variable,
           y = my.y.variable)) +
  geom_point()
```


### Building plots

* For example, consider a scatter plot of daily maximum temperature varying with solar radiation in New York City 1973
* Each row in data has a pair of values $(x,y)$, shown as a point

<!-- * Can show more structure in the data by setting aesthetics of the geometry (mapping variables to graphical elements) -->


``` {r, eval=T, echo=T, fig.width=6, fig.height=2, size = "scriptsize"}
data(airquality)
solar_temp_plot <- ggplot(data = airquality,
       aes(x = Solar.R, y = Temp)) +
  geom_point() 

solar_temp_plot
```

### Scatter plot

* We can add some human-friendly labels and change the theme

\scriptsize

``` {r, eval=T, echo=T, fig.width=6, fig.height=2, size = "small"}
solar_temp_plot <- solar_temp_plot +
  labs(x = 'Solar radiation (Langleys)',
       y = 'Max. daily temp. (F)') +
  theme_bw()

solar_temp_plot
```



<!-- ### Building a plot -- an example -->

<!-- * `bone` data set from `ElemStatLearn` package has `age` of participant and `spnbmd` is rate of change in spinal bone mineral density, two successive measurements divided by time between them -->

<!-- \tiny -->

<!-- ``` {r, eval=T, echo=T, fig.width=4, fig.height=2} -->
<!-- library(ElemStatLearn) -->
<!-- library(ggplot2) -->

<!-- ggplot(data = bone, aes(x = age, y = spnbmd)) + geom_point() -->
<!-- ``` -->


<!-- ### Building a plot -- an example -->

<!-- * This should probably have some meaningful axis labels and we can use a nicer looking theme -->

<!-- \tiny -->

<!-- ``` {r, eval=T, echo=T, fig.width=4, fig.height=2} -->
<!-- ggplot(data = bone, -->
<!--        aes(x = age, y = spnbmd)) + -->
<!--   geom_point() + -->
<!--   labs(x = 'Age (years)', -->
<!--     y = expression(Rel.~BMD~(g~ cm^{-3}~yr^{-1}))) + -->
<!--   theme_bw() -->
<!-- ``` -->


<!-- ### Building a plot -- an example -->

<!-- ``` {r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=4} -->
<!-- library(dplyr) -->
<!-- library(stringi) -->
<!-- library(magrittr) -->

<!-- bone %<>% -->
<!--   count(idnum) %>% -->
<!--   inner_join(bone) %>% -->
<!--   mutate(gender = stri_trans_totitle(gender)) -->

<!-- bone.singleton <- bone %>% -->
<!--   filter(n==1) -->

<!-- bone.repeats <- bone %>% -->
<!--   group_by(idnum) %>% -->
<!--   filter(n > 1) -->

<!-- ggplot(data = bone, -->
<!--        aes(x = age, y = spnbmd)) + -->
<!--   geom_point(alpha = 0.125, stroke=0, pch=19, -->
<!--              size=0.75) + -->
<!--   geom_line(data=bone.repeats, -->
<!--             aes(group = idnum), alpha = 0.05) + -->
<!--   facet_wrap(~ gender) + -->
<!--   theme_bw() + -->
<!--   geom_line(data=bone, stat='smooth', -->
<!--               alpha = 0.25, -->
<!--               #method='gam', formula = y ~s(x, bs='ps', k=6), fill='lightskyblue') + -->
<!--               method='loess', color='blue', -->
<!--             size=1.5) + -->
<!--   labs( -->
<!--     x = 'Age (years)', -->
<!--     y = expression(atop(Rel.~Bone~Mineral, Density~(g~cm^{-3}~yr^{-1}))), -->
<!--     title = 'Relative spinal bone mineral density'#, -->
<!--     #caption = 'Data sourced from Bachrach et al. (1999)\n J Clin Endocrin Metab 84 (12): 4702-12' -->
<!--   ) -->

<!-- ``` -->

<!-- We can keep adding more elements and options to show more details of the data \citep{layered-grammar, cheatsheets} -->


<!-- ### Presentation of plots -->

<!-- ``` {r, eval=T, echo=F, warning=FALSE, message=FALSE, fig.width=4, fig.height=2, fig.cap='Average rate of change (relative) of spinal bone mineral density over time in healthy male and female youths. Lines show individuals with repeated observations. Data sourced from Bachrach et al. (1999) J Clin Endocrin Metab 84 (12): 4702-12', res=300} -->

<!-- #png(width = 6, height = 3, filename = 'fancy_2.png', units = 'in', res = 300) -->
<!-- ggplot(data = bone, -->
<!--        aes(x = age, y = spnbmd)) + -->
<!--   geom_point(alpha = 0.125, stroke=0, pch=19, -->
<!--              size=0.75) + -->
<!--   geom_line(data=bone.repeats, -->
<!--             aes(group = idnum), alpha = 0.05) + -->
<!--   facet_wrap(~ gender) + -->
<!--   theme_bw() + -->
<!--   geom_line(data=bone, stat='smooth', -->
<!--               alpha = 0.25, -->
<!--               #method='gam', formula = y ~s(x, bs='ps', k=6), fill='lightskyblue') + -->
<!--               method='loess', color='blue', -->
<!--             size=1.5) + -->
<!--   labs( -->
<!--     x = 'Age (years)', -->
<!--     #y = expression(Rel.~Bone~Mineral\\n~Density~(g ~ cm ^ {-3}~yr^{-1}))) -->
<!--     y = expression(atop(Rel.~Bone~Mineral, Density~(g~cm^{-3}~yr^{-1})))) -->

<!-- ``` -->

# Some more geometries


### Line plot

* Similar to scatter plot, but joins pairs of values
* Useful when showing how something changes over time
* Use only when $(x,y)$ are ordered pairs of numeric values, e.g. $x$ is time or date
* For this reason, often referred to as **time series plot**

### Line plot

* Show the Ozone concentrations over time

\tiny

``` {r, fig.height=3, fig.width=8, echo=T}
# make the date column
airquality <- mutate(airquality,
                     Date = as.Date(paste('1973', Month, Day, sep = '-')))

airquality_plot <- ggplot(data=airquality, aes(x=Date, y=Ozone)) +
  geom_line() + theme_bw() +
  labs(y='Ozone concentration (ppb)',
       title = 'Daily mean Ozone in NYC (1973)')

```

### Line plot

* If we have an observation whose neighbours are both `NA` values it can't be plotted with a line
* Can layer multiple geometries for same aesthetic mapping

``` {r, fig.height=3, fig.width=8, echo=T, size = "tiny"}
airquality_plot + geom_point()
```


### Smooth plot

* Often too much data in a scatter plot to see pattern
* Maybe we want to show the reader the trend in the data
* `geom_smooth()` generates a **scatterplot smoother** that shows the overall relationship between $y$ and $x$


``` {r, echo=T, message=F, fig.width=6, fig.height=3, size = "small"}
airquality_plot + geom_point() + geom_smooth() 
```


### Boxplot

\tiny

``` {r, fig.height=2, fig.width=4, eval=T, echo=T}

ggplot(data = airquality, aes(x = factor(Month), y = Ozone)) +
  geom_boxplot() + theme_bw() +
  labs(y = 'Ozone conc. (ppb)', x = 'Month')
```

\normalsize

* **outliers** shown as dots, indicating they're far away from typical values

### Histograms

* univariate graphical summary needs only one aesthetic, `x`
* e.g. a histogram of Ozone concentrations


``` {r, eval=T, echo=T, message=F, fig.width=6, fig.height=2, size = "scriptsize"}
ozone_hist <-
  ggplot(data = airquality, aes(x = Ozone)) +  
  geom_histogram(binwidth = 10, boundary = 0) +  
  labs(x = 'Ozone concentration (ppb)') +
  theme_bw()

ozone_hist
```

# Aesthetics

### Aesthetics

* We've seen the $x$ and $y$ positions so far
* We can also map the following options to variables 
    - **size** of point or thickness of boundary
    - **shape** of points
    - **colour** of boundary
    - **alpha** transparency
    - **fill** colour of internals of geometry
    - **group** to repeat geometry for each subset of the data corresponding to the grouping variable
    
* We can also put these (except group) *outside* `aes()` to fix the value for all parts of that geometry
* Any aesthetics specified inside `ggplot()` will be inherited by all geometries for that plot

### Aesthetics

\tiny

``` {r, eval=T, echo=T, fig.height=2, fig.width=4, message=F}
ggplot(data = airquality, aes(x = Ozone)) +  
  geom_histogram(aes(fill = factor(Month)),
                 binwidth = 10, boundary = 0,
                 color = 'black') +  
  labs(x = 'Ozone concentration (ppb)') +
  theme_bw() + 
  scale_fill_viridis_d(name = 'Month')
```

# Small multiples

### Small multiples

* Group a plot by some categorical variable
* Repeat a basic graph for groups in the data
    * air quality data has information about, e.g. months
* Can view 3-5 dimensions in the data on a 2D page
    - Often a better alternative to 3D, since it doesn't distort comparisons
    - Inner axes relate to the smallest X-Y plots
    - Outer axes relate to the grouping variables
* Avoids using loops

### Small multiples

* By adding one extra command we can tell R to repeat the histogram plot of Ozone concentration for each value of `Month`

``` {r, eval=T, echo=T, fig.height=1.5, fig.width=6, message=F}
ozone_hist + facet_wrap( ~ Month, nrow = 1)
```

### Small multiples

* If we have two (or more) grouping variables we can use `facet_grid()` to tell R to repeat the aesthetic and geometries for specified `rows` and `columns` variables


``` {r, echo=TRUE, fig.width = 8, fig.height = 2, out.width = '100%', size = "tiny"}
library(mosaicData)
data(Weather)

Weather_summarised <-
  aggregate(data = Weather, FUN = mean, 
            cbind(high_temp, avg_temp, low_temp) ~ city + month + year)

```

Or using the tidyverse approach

``` {r, eval=FALSE, out.width = '100%', size = "tiny", echo = TRUE}
Weather_summarised <-
  group_by(Weather, city, month, year) %>%
  summarise_at(.vars = vars(contains("temp")),
               .funs = mean)
```

### Small multiples

``` {r, echo=TRUE, fig.width = 8, fig.height = 2, out.width = '100%', size = "scriptsize"}

Weather_plot <-
  ggplot(data = Weather_summarised, 
         aes(x = month, y = avg_temp)) +
  geom_line() +
  geom_ribbon(aes(ymin = low_temp, ymax = high_temp),
              alpha = 0.25) +
  theme_bw() +
  xlab('Month') + ylab(expression(Temperature~(degree*F))) +
  ggtitle(
    'Monthly avg. of daily mean, min and max temperatures') +
  facet_grid(rows = vars(year), cols = vars(city)) +
  scale_x_continuous(breaks = seq(1, 12, by = 3), 
                     minor_breaks = seq(1, 12), 
                     labels = function(x){month.abb[x]})
```

### Small multiples



``` {r, fig.height=3, fig.width=6, eval=T, echo=F, out.width='100%'}
Weather_plot
```





# Summary

### Summary

* We make graphs to tell a story with data
* Should draw reader in and explain what they're seeing
* Plots are built from
    - geometric objects
    - axis scales
    - coordinate systems (linear or logarithmic scale, 2D, 3D, etc.)
    - annotations (e.g. heading in small multiples)

    
### Summary

* Successively building a plot with a grammar of graphics allows development of complex plots from simple elements and small changes \citep{layered-grammar, cheatsheets}
* Choose a plotting geometry that helps tell the story
* Meaningful labels remove ambiguity and confusion
* Be careful not to put too much in

### 

\tiny

``` {r, echo=T, message=F, fig.width=8, fig.height=4, out.width='100%'}
ggplot(data=mtcars, aes(x=wt, y=mpg)) +
  geom_point(alpha=0.75, aes(shape = factor(am, levels = c(0,1),
                                            labels = c('Auto', 'Manual')),
                             color = factor(cyl),
                             size  = factor(gear))) + theme_bw() +
  scale_shape(name = 'Transmission') +
  scale_size_discrete(name = 'Gears') +
  scale_color_brewer(name='Cylinders', palette = 'Pastel2') +
  xlab('Weight (1000lbs)') + ylab('Fuel efficiency (mpg)') +
  geom_smooth(method = 'loess', se=FALSE, color = 'black') +
  geom_smooth(method = 'lm',    se=FALSE, aes(color = factor(cyl)))
```


### Further reading

<!-- * Extra notes on Tufte's principles -->
* History of visualisation
    - \citet{Friendly:05:gfkl}
    - \citet{Friendly:06:hbook}
* Visualisation to help decision making
    - @visstatthink1997
* ggplot2 resources
    - @layered-grammar
    - @cheatsheets
    - @changgraphics

### References
