---
title: "Visualisation with ggplot2"
subtitle: "2031 - Statistical Computing"
author: "Sam Clifford"
date: "2019-11-08"
output:
  beamer_presentation:
    keep_tex: true
    theme: "Hannover"
    colortheme: "dove"
    fonttheme: "professionalfonts"
    latex_engine: xelatex
    citation_package: natbib
    includes: 
      in_header: matrix.tex
header-includes:
 - \usepackage{hyperref}
 - \usepackage{longtable}
bibliography: extras.bib
#csl: chicago-author-date.csl
---
  
```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(ggthemes)

output <- opts_knit$get("rmarkdown.pandoc.to")
knitr::opts_chunk$set(echo = FALSE, collapse=T, warning=FALSE, message=FALSE,
                      fig.align="center", out.width = "100%",
                      fig.width=4,  fig.height=2)
```

# Principles of graphical excellence

### Why do we visualise?

* We don't just make graphs because it's fun (but it totally is!)
* We do it to communicate information
* Describing with summary statistics may not tell the whole story
* Graph must communicate what's in your mind to reader, including key relationships
* Reader should be able to understand what the graph means and not be
    - misled into thinking something that is untrue
    - distracted from the main point


### Tufte's principles

\citet{tufte} and \citet{pantoliano}

* Show the data
    - View should think about substance of the findings rather than methodology, graphical design, or other aspects
    - Avoid distorting what the data have to say
    - Serve a clear purpose:  description, exploration, tabulation, or decoration

* Provide clarity
    - Present many numbers in a small space (as opposed to a table)
    - Make large data sets coherent
    - Be closely integrated with the statistical and verbal descriptions of the data set
  
### Tufte's principles

* Allow comparison where appropriate
    - Encourage the eye to compare different pieces of data
    - Reveal the data at several levels of detail, from a broad overview to the fine structure

* Produce graphs with high data density by
    * maximising data-ink and erasing non-data-ink 
    * careful use of whitespace
* Avoid excessive and unnecessary use of graphical effects (chartjunk)
* Don't map the same variable to multiple graphical elements (e.g. color and $y$ value)

    
# Building plots

### Building plots

* R package `ggplot2` uses a grammar of graphics [@layered-grammar; @cheatsheets]
    - map variables in data frame to aesthetic options in the plot
    - choose a geometry for how to display these variables
    - adjustments to axis scales
    - adjustments to colors, themes, etc.
    - adding extra commands in a "do this, then do this" manner
  
### Building a plot

* How do we structure a call to `ggplot` to make a plot?
    - load `ggplot2` package
    - Specify we want a `ggplot` object and which data frame we're going to use, 
    - set **aesthetic options** to tell R which variables to map to the $x$ and $y$ axes of the plot
    - state geometry we're using to show variables
    
``` {r, eval=F, echo=T}
ggplot(data = my.data.frame, 
       aes(x = my.x.variable,
           y = my.y.variable)) +
  geom_point()
```

### Building a plot -- an example

* `bone` data set from `ElemStatLearn` package has `age` of participant and `spnbmd` is rate of change in spinal bone mineral density, two successive measurements divided by time between them

\tiny

``` {r, eval=T, echo=T, fig.width=4, fig.height=2}
library(ElemStatLearn)
library(ggplot2)

ggplot(data = bone, aes(x = age, y = spnbmd)) + geom_point()
```


### Building a plot -- an example

* This should probably have some meaningful axis labels and we can use a nicer looking theme

\tiny

``` {r, eval=T, echo=T, fig.width=4, fig.height=2}
ggplot(data = bone,
       aes(x = age, y = spnbmd)) +
  geom_point() +
  labs(x = "Age (years)",
    y = expression(Rel.~BMD~(g~ cm^{-3}~yr^{-1}))) +
  theme_bw()
```


### Building a plot -- an example

``` {r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=2, fig.width=4}
library(dplyr)
library(stringi)
library(magrittr)

bone %<>%
  count(idnum) %>%
  inner_join(bone) %>%
  mutate(gender = stri_trans_totitle(gender))

bone.singleton <- bone %>%
  filter(n==1)

bone.repeats <- bone %>%
  group_by(idnum) %>%
  filter(n > 1)

ggplot(data = bone,
       aes(x = age, y = spnbmd)) +
  geom_point(alpha = 0.125, stroke=0, pch=19,
             size=0.75) +
  geom_line(data=bone.repeats,
            aes(group = idnum), alpha = 0.05) +
  facet_wrap(~ gender) +
  theme_bw() +
  geom_line(data=bone, stat="smooth",
              alpha = 0.25,
              #method="gam", formula = y ~s(x, bs="ps", k=6), fill="lightskyblue") +
              method="loess", color="blue",
            size=1.5) +
  labs(
    x = "Age (years)",
    y = expression(atop(Rel.~Bone~Mineral, Density~(g~cm^{-3}~yr^{-1}))),
    title = "Relative spinal bone mineral density"#,
    #caption = "Data sourced from Bachrach et al. (1999)\n J Clin Endocrin Metab 84 (12): 4702-12"
  )

```

We can keep adding more elements and options to show more details of the data \citep{layered-grammar, cheatsheets}

<!-- ### Presentation of plots -->

<!-- ``` {r, eval=T, echo=F, warning=FALSE, message=FALSE, fig.width=4, fig.height=2, fig.cap="Average rate of change (relative) of spinal bone mineral density over time in healthy male and female youths. Lines show individuals with repeated observations. Data sourced from Bachrach et al. (1999) J Clin Endocrin Metab 84 (12): 4702-12", res=300} -->

<!-- #png(width = 6, height = 3, filename = "fancy_2.png", units = "in", res = 300) -->
<!-- ggplot(data = bone, -->
<!--        aes(x = age, y = spnbmd)) + -->
<!--   geom_point(alpha = 0.125, stroke=0, pch=19, -->
<!--              size=0.75) + -->
<!--   geom_line(data=bone.repeats, -->
<!--             aes(group = idnum), alpha = 0.05) + -->
<!--   facet_wrap(~ gender) + -->
<!--   theme_bw() + -->
<!--   geom_line(data=bone, stat="smooth", -->
<!--               alpha = 0.25, -->
<!--               #method="gam", formula = y ~s(x, bs="ps", k=6), fill="lightskyblue") + -->
<!--               method="loess", color="blue", -->
<!--             size=1.5) + -->
<!--   labs( -->
<!--     x = "Age (years)", -->
<!--     #y = expression(Rel.~Bone~Mineral\\n~Density~(g ~ cm ^ {-3}~yr^{-1}))) -->
<!--     y = expression(atop(Rel.~Bone~Mineral, Density~(g~cm^{-3}~yr^{-1})))) -->

<!-- ``` -->

# Some common geometries


### Scatter plot

* Each row in data has a pair of values $(x,y)$, shown as a point
* Can show more structure in the data by setting aesthetics of the geometry (mapping variables to graphical elements)

\footnotesize

``` {r, eval=F, echo=T, fig.width=4, fig.height=2}
data(airquality)
ggplot(data = airquality,
       aes(x = Solar.R, y = Temp)) +
  geom_point() +
  labs(x = "Solar radiation (Langleys)",
       y = "Max. daily temperature (F)") +
  theme_bw()
```

### Scatter plot


``` {r, eval=T, echo=F, fig.width=4, fig.height=2}
ggplot(data = airquality,
       aes(x = Solar.R, y = Temp)) +
  geom_point() +
  labs(x = "Solar radiation (Langleys)",
       y = "Max. daily temperature (F)") +
  theme_bw()
```



### Line plot

* Similar to scatter plot, but joins pairs of values
* Useful when showing how something changes over time
* Use only when $(x,y)$ are ordered pairs of numeric values, e.g. $x$ is time or date
* For this reason, often referred to as **time series plot**

### Line plot

* Show the Ozone concentrations over time

\tiny

``` {r, fig.height=1.75, fig.width=4, echo=T}
# make the date column
airquality$Date <- as.Date(paste("1973", 
                                 airquality$Month,
                                 airquality$Day,
                                 sep="-"))

ggplot(data=airquality, aes(x=Date, y=Ozone)) +
  geom_line() + theme_bw() +
  labs(y="Concentration (ppb)",
       title = "Daily mean Ozone in NYC (1973)")
```

### Line plot

* `geom_line()` stops plotting when it hits an `NA` value
* If we have individual measurements in a group of `NA` values it won't plot that value 
* Can use multiple geometries to display the same variables

\tiny

``` {r, fig.height=1.75, fig.width=4, echo=T}
ggplot(data=airquality, aes(x=Date, y=Ozone)) +
  geom_line() + geom_point() + theme_bw() +
  labs(y="Concentration (ppb)",
       title = "Daily mean Ozone in NYC (1973)")
```


### Smooth plot

* Often too much data in a scatter plot to see pattern
* Maybe we want to show the reader the trend in the data
* `geom_smooth()` generates a **scatterplot smoother** that shows the overall relationship between $y$ and $x$

\tiny 

``` {r, echo=T, message=F, fig.width=4, fig.height=2}
ggplot(data=airquality, aes(x=Temp, y=Ozone)) +
  geom_point() + geom_smooth() + theme_bw() +
  labs(x=expression(Temperature~(degree*F)), y="Ozone (ppb)")
```


### Boxplot

\tiny

``` {r, fig.height=2, fig.width=4, eval=T, echo=T}

ggplot(data = airquality, aes(x = factor(Month), y = Ozone)) +
  geom_boxplot() + theme_bw() +
  labs(y = "Ozone conc. (ppb)", x = "Month")
```

\normalsize

* **outliers** shown as dots, indicating they're far away from typical values

### Bar/column plots

* Good for showing the amount of something (requires meaningful zero)
* `geom_col()` - $xy$ plot given some variables `x`, `y`
* `geom_bar()` - counts number of of times categorical $x$ occurs
* `geom_histogram()` - counts number of times $x$ in bin

### Bar/column plots

* e.g. a histogram of Ozone concentrations

\tiny 

``` {r, eval=T, echo=T, message=F, fig.width=4, fig.height=2}
ozone_hist <-
  ggplot(data = airquality, aes(x = Ozone)) +  
  geom_histogram(binwidth = 10, boundary = 0) +  
  labs(x = "Ozone concentration (ppb)") +
  theme_bw()

ozone_hist
```

### Bar/column plots

* e.g. a bar plot of month

\footnotesize

``` {r, eval=T, echo=T, message=F, fig.width=4, fig.height=2, out.width = "100%"}
ggplot(data = airquality, aes(x = factor(Month))) +  
  geom_bar() + labs(x = "Month") + theme_bw()

```


# Small multiples

### Small multiples

* Group a plot by some categorical variable
* Repeat a basic graph for groups in the data
    * air quality data has information about, e.g. months
* Can view 3-5 dimensions in the data on a 2D page
    - Often a better alternative to 3D, since it doesn't distort comparisons
    - Inner axes relate to the smallest X-Y plots
    - Outer axes relate to the grouping variables
* Avoids using loops

### Small multiples

* By adding one extra command we can tell R to repeat the histogram plot of Ozone concentration for each value of `Month`

``` {r, eval=T, echo=T, fig.height=1.5, fig.width=6, message=F}
ozone_hist + facet_wrap( ~ Month, nrow = 1)
```

### Small multiples

* If we have two (or more) grouping variables we can use `facet_grid()` to tell R to repeat the aesthetic and geometries for specified `rows` and `columns` variables



\tiny



``` {r, echo=TRUE, fig.width = 8, fig.height = 2, out.width = "100%"}
library(mosaicData)
library(data.table)

data(Weather)

Weather <- data.table(Weather)
  
Weather_summarised <-
  Weather[ , 
           lapply(.SD, mean),
           by = .(city, month, year),
           .SDcols = patterns("temp")]

Weather_plot <-
  ggplot(data = Weather_summarised, aes(x = month, y = avg_temp)) +
  geom_line() +
  geom_ribbon(aes(ymin = low_temp, ymax = high_temp), alpha = 0.25) +
  theme_bw() +
  xlab("Month") + ylab(expression(Temperature~(degree*F))) +
  ggtitle("Monthly average of daily mean, min and max temperatures") +
  facet_grid(rows = vars(year), cols = vars(city)) +
  scale_x_continuous(breaks = seq(1, 12, by = 3))

```


### Small multiples



``` {r, fig.height=3, fig.width=6, eval=T, echo=F, out.width="100%"}
Weather_plot
```



### Changing the default options

\tiny

``` {r, echo=T, message=F, fig.width=6, fig.height=3, out.width="100%"}
ggplot(data=mtcars, aes(x=wt, y=mpg)) +
  geom_point(alpha=0.75, aes(shape=factor(am), color = factor(cyl),
                             size=factor(gear))) + theme_bw() +
  scale_shape(name="Transmission", breaks=c("0", "1"),
              labels=c("Auto", "Manual")) +
  scale_size_discrete(name="Gears") +
  scale_color_brewer(name="Cylinders", palette = "Pastel2") +
  xlab("Weight (1000lbs)") + ylab("Fuel efficiency (mpg)") +
  geom_smooth(method = "loess", se=FALSE, color = "black")
```


# Summary

### Summary

* We make graphs to tell a story with data
* Graphs should draw the reader in and explain what they're seeing
* Plots are built from
    - geometric objects
    - axis scales
    - coordinate systems (linear scale, logarithmic scale, 2D, 3D, etc.)
    - annotations (e.g. heading in small multiples)

    
### Summary

* Successively building a plot with a grammar of graphics allows development of complex plots from simple elements and small changes
* Choose a plotting geometry that helps tell the story
* Meaningful labels remove ambiguity and confusion

### Further reading

* Extra notes on Tufte's principles
* History of visualisation
    - @Friendly:05:gfkl
    - @Friendly:06:hbook
* Visualisation to help decision making
    - @visstatthink1997
* ggplot2 resources
    - @layered-grammar
    - @cheatsheets

\scriptsize
